---
title: "1. Getting started with STEM models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{1. Getting started with STEM models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6.5,
  fig.height = 4
)
```

```{r load}
library(Stem)
```

## The model

`Stem` estimates hierarchical spatio-temporal models of the form

$$
z_t = X_t \beta + K y_t + e_t, \qquad e_t \sim N(0, \Sigma_e),
$$
$$
y_t = G y_{t-1} + \eta_t, \qquad \eta_t \sim N(0, \Sigma_\eta), \qquad
y_0 \sim N(m_0, C_0),
$$

for $t = 1, \dots, T$. The observation vector $z_t$ collects the measurements at
$d$ spatial locations; $y_t$ is a $p$-dimensional latent temporal process shared
by the network and loaded on the locations through the $d \times p$ matrix $K$;
$X_t$ holds $r$ known covariates. The spatial covariance is

$$
\Sigma_e = \sigma^2_\varepsilon I + \sigma^2_\omega C(h; \theta), \qquad
C(h; \theta) = \exp(-\theta h),
$$

so that $\sigma^2_\varepsilon$ is a nugget and $\sigma^2_\omega$ the variance of
the spatially correlated component. The parameter vector is
$\phi = (\beta, \sigma^2_\varepsilon, \sigma^2_\omega, \theta, G, \Sigma_\eta,
m_0)$, estimated by maximum likelihood through Kalman filtering and smoothing
inside an EM algorithm, with $C_0$ held fixed.

## Data

The package ships two datasets. `pm10` is the small example of Fasso and
Cameletti: 22 stations and 366 days of log PM10, with emissions and altitude as
covariates. `povalley` is a larger, real network of 36 background stations in
the Po Valley observed daily over 2019--2023.

```{r data}
data(pm10)

str(pm10, max.level = 1)

coordinates <- pm10$coords * 1000   # from km to meters
covariates <- pm10$covariates
z <- pm10$z

dim(z)            # T x d
dim(covariates)   # (T * d) x r
```

The covariate matrix stacks the $T \times r$ block of each station on top of the
next: rows $1, \dots, T$ belong to the first station, rows $T+1, \dots, 2T$ to
the second, and so on. Keeping this convention is essential, because every
routine of the package relies on it to slice the data by location.

## Building a model object

`STEM_Model()` assembles the data and the starting values into an object of
class `STEM_Model`. The starting values are supplied through the list `phi`,
which must have exactly the components of the parameter vector.

```{r model}
phi <- list(
  beta        = matrix(c(3.65, 0.046, -0.904), 3, 1),
  sigma2eps   = 0.1,
  sigma2omega = 0.2,
  theta       = 0.01,
  G           = matrix(0.77, 1, 1),
  Sigmaeta    = matrix(0.3, 1, 1),
  m0          = as.matrix(0),
  C0          = as.matrix(1)
)

K <- matrix(1, ncol(z), 1)   # a single latent process, loaded on every station

mod <- STEM_Model(
  z = z,
  covariates = covariates,
  coordinates = coordinates,
  phi = phi,
  K = K
)

class(mod)
```

## Estimation

`STEM_Estimation()` runs the EM algorithm. Convergence is declared when both the
relative change of the parameter vector and the relative change of the
log-likelihood fall below `precision`. Progress messages are off by default and
can be turned on with `verbose = TRUE`.

```{r estimation}
fit <- STEM_Estimation(mod, precision = 0.01, distance = "euclidean")

round(unlist(fit$estimates$phi.hat), 5)

fit$estimates$loglik
fit$estimates$convergence.par[c("conv.log", "conv.par", "iterEM")]
```

Use `distance = "geo"` when the coordinates are longitude and latitude, as in
`povalley`; the geodesic distance is then computed with `geodist`.

The smoothed latent process is returned as a `ts` object:

```{r smoothed, fig.alt = "Smoothed latent temporal process"}
plot(fit$estimates$y.smoothed, ylab = "smoothed latent state", xlab = "day")
```

## Uncertainty: the parametric bootstrap

The EM algorithm returns point estimates but no standard errors.
`STEM_Bootstrap()` simulates `B` datasets from the fitted model and re-estimates
the parameters on each of them.

```{r bootstrap, eval = FALSE}
boot <- STEM_Bootstrap(fit, B = 100, distance = "euclidean")

npar <- 9
boot.estimates <- t(vapply(boot$boot.output, function(b) {
  p <- b$estimates$phi.hat
  c(p$beta, p$sigma2eps, p$sigma2omega, p$theta, p$G, p$Sigmaeta, p$m0)
}, numeric(npar)))

se <- sqrt(diag(stats::var(stats::na.omit(boot.estimates))))
round(se, 5)
```

## Spatial prediction

`STEM_Kriging()` predicts the process at unobserved locations, given their
coordinates, covariates and loading matrix.

## Where to go next

The second vignette, *Spatially-clustered STEM models*, shows what to do when a
single global regression is not plausible over the whole network, and the
coefficients themselves should be allowed to vary across spatial regimes:

```{r next, eval = FALSE}
vignette("SCSTEM", package = "Stem")
```

A conceptual map of how the functions call one another is the third vignette:

```{r map, eval = FALSE}
vignette("function-map", package = "Stem")
```

## References

Fasso, A., Cameletti, M. and Nicolis, O. (2007). Air quality monitoring using
heterogeneous networks. *Environmetrics*, 18, 245--264.

Fasso, A. and Cameletti, M. (2010). A unified statistical approach for
simulation, modeling, analysis and mapping of environmental data. *Simulation*,
86, 139--153.
