---
title: "Introduction to the STEM package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to the STEM package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```


# Overview

`Stem` implements hierarchical spatio-temporal models estimated through the Expectation-Maximization (EM) algorithm. The package is intended for panel data observed at multiple spatial locations over time, such as environmental monitoring data, where latent temporal dynamics and spatial covariance are both relevant.

The package also provides spatially-clustered STEM routines. These functions combine clustering with iterative parameter estimation, allowing groups of spatial units to share local regression dynamics while preserving the broader STEM modelling structure.

# The STEM workflow

A typical analysis has three steps:

1. organize observations, covariates, coordinates, and starting values into a `STEM_Model` object;
2. estimate model parameters with `STEM_Estimation`;
3. use the fitted model for simulation, bootstrap uncertainty assessment, or spatial prediction.

# Example: basic STEM model

```{r basic-stem, eval = FALSE}
library(Stem)
data(pm10)

coordinates <- pm10$coords * 1000
covariates <- pm10$covariates
z <- pm10$z

phi <- list(
  beta = matrix(c(3.65, 0.046, -0.904), 3, 1),
  sigma2eps = 0.1,
  sigma2omega = 0.2,
  theta = 0.01,
  G = matrix(0.77, 1, 1),
  Sigmaeta = matrix(0.3, 1, 1),
  m0 = as.matrix(0),
  C0 = as.matrix(1)
)

K <- matrix(1, ncol(z), 1)

mod <- STEM_Model(
  z = z,
  covariates = covariates,
  coordinates = coordinates,
  phi = phi,
  K = K
)

fit <- STEM_Estimation(mod, distance = "euclidean")
fit$estimates$phi.hat
```

The chunks are not evaluated during vignette building because the example estimation can be computationally demanding on CRAN machines. They are intended as executable templates for users.

# Example: spatially-clustered STEM

```{r scstem, eval = FALSE}
sc_fit <- SCSTEM_Estim(
  StemModel = mod,
  distance = "euclidean",
  crs = 32632,
  k = 3,
  verbose = TRUE
)

head(sc_fit$df)
```

The function `SCSTEM_Infocrit` can be used to compare a range of candidate cluster numbers through likelihood-based information criteria.

```{r scstem-ic, eval = FALSE}
ic_tab <- SCSTEM_Infocrit(
  StemModel = mod,
  mink = 2,
  maxk = 4,
  distance = "euclidean",
  crs = 32632,
  verbose = TRUE
)

ic_tab
```

# Function map

The package includes an updated call graph in `inst/extdata/STEM_function_map.svg`, generated by statically parsing calls among package functions. The original conceptual map is retained in `inst/extdata/STEM_function_map_original.pdf`.

# References

Fassò, Cameletti and Nicolis (2007) and Fassò and Cameletti (2010) describe the original STEM modelling framework. Sugasawa and Murakami (2021) provide the spatially-clustered regression approach that motivates the clustered extension.
